---
title: "forcats"
author: "Janine De Vera and Victor Möslein"
date: "2021 November 04"
output:
  ioslides_presentation: default
  incremental: true
---

<style type = "text/css">
h2 {color: #0077b7;}
</style>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r packages, include=FALSE}
pacman::p_load(tidyverse, knitr, kableExtra)
```

```{r theme, include=FALSE}
charts_theme <- theme(axis.line.x = element_line("transparent", size = 0.5), 
                      axis.line.y = element_line("transparent", size = 0.5),
                      axis.title.y.left = element_text(size = 12, margin = margin(r = 15)),
                      axis.title.y.right = element_text(size = 12, margin = margin(l = 15)),
                      axis.title.x = element_text(size = 12, margin = margin(t = 15)),
                      panel.border = element_rect(color = "#a3a3a3", fill = "transparent"),
                      panel.background = element_rect(fill = "white", color = "white"),
                      panel.grid.major = element_line(color = "#d4d4d4", linetype = 2),
                      plot.title = element_text(size = 14, hjust = 0.5),
                      plot.subtitle = element_text(size = 12, face = "italic", hjust = 0.5, margin = margin(b = 15)),
                      legend.background = element_rect(fill = "transparent"),
                      legend.text = element_text(size = 12),
                      legend.title = element_blank(),
                      legend.key = element_blank())
```

## Outline
1. **Introduction:** data manipulation with `forcats`
2. **Features:** common uses of `forcats`
    - Reordering factors
    - Renaming factors 
3. **Exercises** 

# Introduction {data-background=#007db7} 

## Dealing with categorical data in R {.build}
_**Recall:**_ \
\
A **categorical variable** is data that takes a limited or finite set of possible values. Assignment of observation to a group or nominal category is often based on a qualitative property. _(e.g. gender, marital status, race)_

## Dealing with categorical data in R {.build}

Categorical variables are stored in R either as _character strings_ or _factors_. \

_**Factors**_ are <u>integer vectors</u> that are mapped to categorical values. Factors contain a `levels` attribute, which is the order at which categorical values are assigned to integers. \

## Dealing with categorical data in R {.build .smaller}
Suppose we have a variable called `rate` that takes on the values **low**, **medium**, and **high**. When stored as character strings, R will sort the values by alphabetical order. 

`rate <- c("low", "medium", "high")` \
`sort(rate)`
```{r}
rate <- c("low", "medium", "high")
sort(rate)
```

When converted into a factor, R will provide additional information on the variable **levels**.

`rate_factor <- as.factor(rate)`
```{r}
rate_factor <- as.factor(rate)
rate_factor
```

## Dealing with categorical data in R {.build .smaller}

To illustrate the mapping of categorical values to integers, we will try to recover numeric information from `rate_factor`: 

<font size = 4> `rate_level <- as.numeric(rate_factor)` </font>
<br>
<br>
```{r}
as.data.frame(rate_factor) %>%
  mutate(rate_level = as.numeric(rate_factor)) %>% 
  kable(col.names = c("rate", "rate_level")) %>% 
  kable_paper() %>% 
  kable_styling(full_width = F, position = "center")
```
\
Note that if we do this for the original vector, we will not be able to extract any numeric information. 

`as.numeric(rate)`
```{r, warning = FALSE}
as.numeric(rate)
```


# Features {data-background=#007db7} 

## Sample {.build .smaller} 
For demonstration of the features of `forcats`, we will be working with the built-in dataset `gss_cat`. 

The **General Social Survey** is a large long-running survey of US citizens with pertinent socio-economic variables. The sample from `gss_cat` includes results from the years 2000-2014. 

Below is a summary of the data:
```{r}
data <- gss_cat 
head(data, n=5)
```

## Sample {.build .smaller} 
Let's examine the factor variable `marital`. \

To access factors levels directly, we can use the function `levels()`. 

`levels(data$marital)`
```{r}
levels(data$marital)
```
\
\
\
Suppose we want to visualize the percentage share of each marital status in the sample.

## Sample
```{r}
m_share <- data %>% 
  count(marital) %>% 
  mutate(share = n/sum(n))

ggplot(m_share, aes(x = marital, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = marital,
                   yend = share,
                   xend = marital), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Marital Status Categories",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("marital status") +
  ylab("") +
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  coord_flip()
```

---
R displays factor information in charts (and tables) based on the variable's defined levels. \

While the chart was already quite informative, we may want to present the order of our variable in another manner.

# Reordering factors {data-background=#68c5ea} 

## A. Reordering (manual) {.build .smaller}

We can reorder factor levels based on our desired arrangement. 

One option is to **reorder factors manually** using the `fct_relevel()` function. \

The following arguments need to be specified: \
1. the factor of interest (as a vector) \
2. the desired factor order \

Recall, the levels of the `marital` variable: 
```{r}
levels(data$marital)
```

## A. Reordering (manual) {.build .smaller}
If we want the category `Married` to be the first level, we can do so by specifying it as the second argument:

`fct_relevel(data$marital, "Married")`
```{r}
data$marital <- fct_relevel(data$marital, "Married")
levels(data$marital)
```

We can also use the argument `after` to specify a position. This places `Widowed` after the 2nd level or in the 3rd place: 

`fct_relevel(data$marital, "Widowed", after = 2)`
```{r}
data$marital <- fct_relevel(data$marital, "Widowed", after = 2)
levels(data$marital)
```

This places `No answer` in the last place: 

`fct_relevel(data$marital, "No answer", after = Inf)`
```{r}
data$marital <- fct_relevel(data$marital, "No answer", after = Inf)
levels(data$marital)
```

## A. Reordering (manual)
```{r}
m_share <- data %>% 
  count(marital) %>% 
  mutate(share = n/sum(n))

ggplot(m_share, aes(x = marital, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = marital,
                   yend = share,
                   xend = marital), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Marital Status Categories",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("marital status") +
  ylab("") +
  scale_y_continuous(labels = function(x) paste0(x*100, "%")) +
  coord_flip()
```

## B. Reordering (variable-based) {.build .smaller}

We can also reorder a factor **based on another variable.**

Suppose we want to plot marital status against another variable, `tvhours`, to determine the average number of hours spent watching TV by each group. 

## B. Reordering (variable-based)
```{r, message=FALSE}
m_tv <- data %>% 
  filter(year %in% c(2000, 2014)) %>% 
  group_by(marital, year) %>% 
  summarise(across(tvhours, mean, na.rm = T))

ggplot(m_tv, aes(x = marital, y = tvhours)) +
  geom_segment(m_tv %>% spread(., year, tvhours),
               mapping = 
                 aes(y = `2000`,
                   x = marital,
                   yend = `2014`,
                   xend = marital), 
               color = "#007db7",
               size = 1.5) +
  geom_point(aes(color = as.character(year)), stat = "identity", size = 6) +
  charts_theme + 
  labs(title = "TV Hours by Marital Status",
       subtitle = "2000 vs 2014") + 
  xlab("marital status") +
  ylab("average tv hours") +
  scale_y_continuous(limits = c(1,5)) +
  scale_color_manual(values = c("#007db7", "#f57f29")) +
  coord_flip()
```

## B. Reordering (variable-based) {.build .smaller}
We can make the chart more readable by reporting the `marital` variable in order of average `tvhours`. \

For variable-based reordering, we use the function `fct_reorder()`. 

The following arguments need to be specified: \
1. the factor of interest (as a vector) \
2. the variable to base the order on \

`levels(data$marital)`
```{r}
levels(m_tv$marital)
```

`fct_reorder(data$marital, data$tvhours)`
```{r}
m_tv$marital <- fct_reorder(m_tv$marital, m_tv$tvhours)
levels(m_tv$marital)
```

## B. Reordering (variable-based)
```{r}
ggplot(m_tv, aes(x = marital, y = tvhours)) +
  geom_segment(m_tv %>% spread(., year, tvhours),
               mapping = 
                 aes(y = `2000`,
                   x = marital,
                   yend = `2014`,
                   xend = marital), 
               color = "#007db7",
               size = 1.5) +
  geom_point(aes(color = as.character(year)), stat = "identity", size = 6) +
  charts_theme + 
  labs(title = "TV Hours by Marital Status",
       subtitle = "2000 vs 2014") + 
  xlab("marital status") +
  ylab("average tv hours") +
  scale_y_continuous(limits = c(1,5)) +
  scale_color_manual(values = c("#007db7", "#f57f29")) +
  coord_flip()

```

## C. Reordering (frequency-based) {.build .smaller}
Now suppose we want to construct a frequency table for marital status and graph the results. \

```{r}
mar <- data %>% 
  filter(year %in% c("2000", "2014")) %>% 
  select(year, marital) %>% 
  group_by(year) %>% 
  count(marital)

mar %>% 
  spread(., year, n) %>% 
  kable(col.names = c("Marital Status", "2000", "2014")) %>% 
  kable_paper() %>% 
  kable_styling(full_width = F, position = "center")
```

## C. Reordering (frequency-based) 
```{r}
ggplot(mar, aes(x = marital, y = n)) +
  geom_bar(aes(fill = as.character(year)), 
           position = "dodge", 
           stat = "identity", 
           width = 0.7) +
  charts_theme + 
  labs(title = "Marital Status Frequencies",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("marital status") +
  ylab("frequency") +
  scale_fill_manual(values = c("#007db7", "#fdb515")) +
  coord_flip()
```

## C. Reordering (frequency-based) {.build .smaller}
We can easily reorder the levels based on how frequent a value appears on the sample using the function `fct_infreq()`.

`levels(data$marital)`
```{r}
levels(data$marital)
```

`fct_infreq(data$marital)`
```{r}
data$marital <- fct_infreq(data$marital)
levels(data$marital)
```

## C. Reordering (frequency-based)
```{r}
mar <- data %>% 
  filter(year %in% c("2000", "2014")) %>% 
  select(year, marital) %>% 
  group_by(year) %>% 
  count(marital)

ggplot(mar, aes(x = marital, y = n)) +
  geom_bar(aes(fill = as.character(year)), 
           position = "dodge", 
           stat = "identity", 
           width = 0.6) +
  charts_theme + 
  labs(title = "Marital Status Frequencies",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("marital status") +
  ylab("frequency") +
  scale_fill_manual(values = c("#007db7", "#fdb515")) +
  coord_flip()
```

## D. Reordering (reverse) {.build .smaller}

Another useful function is reversing the order of levels using `fct_rev()`.

`levels(data$marital)`
```{r}
levels(data$marital)
```

`fct_rev(data$marital)`
```{r}
levels(fct_rev(data$marital))
```

## D. Reordering (reverse)
```{r}
ggplot(mar, aes(x = fct_rev(marital), y = n)) +
  geom_bar(aes(fill = as.character(year)), 
           position = "dodge", 
           stat = "identity", 
           width = 0.6) +
  charts_theme + 
  labs(title = "Marital Status Frequencies",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("marital status") +
  ylab("frequency") +
  scale_fill_manual(values = c("#007db7", "#fdb515")) +
  coord_flip()
```

# Renaming factors {data-background=#68c5ea} 

## Renaming factors {.build .smaller}
Another useful feature of `forcats` is changing level **values**. Doing so makes labels clearer for data visualization and also allows us to collapse and combine levels.

Take for example the `partyid` variable which shows the respondent's party affiliation. 

```{r}
party_share <- data %>% 
               count(partyid) %>% 
               mutate(share = round(n/sum(n)*100, 2)) %>% 
               select(!n)

party_share %>% 
  kable(col.names = c("Party", "% Share")) %>% 
  kable_paper() %>% 
  kable_styling(full_width = F, position = "center")
```

## Renaming factors
```{r}
ggplot(party_share, aes(x = partyid, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = partyid,
                   yend = share,
                   xend = partyid), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Political Parties",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("party affilation") +
  ylab("") +
  scale_y_continuous(limits = c(0,25),
    labels = function(x) paste0(x, "%")) +
  coord_flip()
```

## A. Renaming (recoding) {.build .smaller}

The `partyid` variable has 10 levels, three of which are not specifically defined - \
_No answer_, _Don't know_, and _Other party_.

What we may want to do is combine these three into a single category. We can use the function `fct_recode()` to change the labels of the factor levels. 

The following arguments need to be specified: \
1. the factor of interest (as a vector) \
2. mapping of new and old factor values \

<font size = 3>
`data %>%` \
`mutate(fct_recode(partyid,` \
`"Other" = "No answer",` \
`"Other" = "Don't know",` \
`"Other" = "Other party"))`
</font>

## A. Renaming (recoding)
```{r}
data <- data %>% 
  mutate(partyid = fct_recode(partyid,
                   "Other" = "No answer",
                   "Other" = "Don't know",
                   "Other" = "Other party"))

party_share <- data %>% 
               count(partyid) %>% 
               mutate(share = round(n/sum(n)*100, 2)) %>% 
               select(!n)

ggplot(party_share, aes(x = partyid, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = partyid,
                   yend = share,
                   xend = partyid), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Political Parties",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("party affilation") +
  ylab("") +
  scale_y_continuous(limits = c(0,25),
    labels = function(x) paste0(x, "%")) +
  coord_flip()
```

## A. Renaming (recoding) {.build .smaller}
We can also use the `fct_recode()` function to clean up and standardize the values of the factors. 

<font size = 3>
`data %>%` 
`mutate(partyid = fct_recode(partyid,` \
    `"Republican, strong"    = "Strong republican",`\
    `"Republican, weak"      = "Not str republican",` \
    `"Independent, near rep" = "Ind,near rep",` \
    `"Independent, near dem" = "Ind,near dem",` \
    `"Democrat, weak"        = "Not str democrat",` \
    `"Democrat, strong"      = "Strong democrat"))` \
</font>

## A. Renaming (recoding)
```{r}
data <- data %>% 
mutate(partyid = fct_recode(partyid, 
    "Republican, strong"    = "Strong republican",
    "Republican, weak"      = "Not str republican",
    "Independent, near rep" = "Ind,near rep",
    "Independent, near dem" = "Ind,near dem",
    "Democrat, weak"        = "Not str democrat",
    "Democrat, strong"      = "Strong democrat"))

party_share <- data %>% 
               count(partyid) %>% 
               mutate(share = round(n/sum(n)*100, 2)) %>% 
               select(!n)

ggplot(party_share, aes(x = partyid, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = partyid,
                   yend = share,
                   xend = partyid), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Political Parties",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("party affilation") +
  ylab("") +
  scale_y_continuous(limits = c(0,25),
    labels = function(x) paste0(x, "%")) +
  coord_flip()
```

## B. Renaming (collapsing) {.build .smaller}
For combining multiple levels together, and even more useful function is `fct_collapse()`. \

The function allows us to map the new value to a vector of old values. 

So if we want to combine all Republicans, Democrats and Independents: 

<font size = 3>
`data %>%` \
`mutate(partyid = fct_collapse(partyid,` \
`Republican = str_subset(data$partyid, "Republican"),` \
`Independent = str_subset(data$partyid, "Independent"),` \
`Democrat = str_subset(data$partyid, "Democrat")))` \
</font>

```{r}
data <- data %>% mutate(partyid = 
           fct_collapse(partyid,
                        Republican = str_subset(data$partyid, "Republican"),
                        Independent = str_subset(data$partyid, "Independent"),
                        Democrat = str_subset(data$partyid, "Democrat")))
```

## B. Renaming (collapsing)
```{r}
party_share <- data %>% 
               count(partyid) %>% 
               mutate(share = round(n/sum(n)*100, 2)) %>% 
               select(!n)

ggplot(party_share, aes(x = partyid, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = partyid,
                   yend = share,
                   xend = partyid), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Political Parties",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("party affilation") +
  ylab("") +
  scale_y_continuous(limits = c(0,50),
    labels = function(x) paste0(x, "%")) +
  coord_flip()
```

## C. Renaming (lumping) {.build .smaller}
If there are no specific values that we want to collapse together, we may also want to just lump the smallest values into a single category. 

To do this we use the `fct_lump()` function. 

The following arguments need to be specified: \
1. the factor of interest (as a vector) \
2. the number of categories that we want to retain \

## C. Renaming (lumping) {.build .smaller}

We have the original table:
<br>
<br>
```{r}
data <- gss_cat

data %>% 
  count(partyid) %>% 
  mutate(share = round(n/sum(n)*100, 2)) %>% 
  select(!n) %>% 
  kable(col.names = c("Party", "% Share")) %>% 
  kable_paper() %>% 
  kable_styling(full_width = F, position = "center")
```

## C. Renaming (lumping) {.build .smaller}
And we want to retain only the top 4 categories and move the rest to the _"Other"_ category. 

`data %>%` 
`mutate(partyid = fct_lump(partyid, n = 4))`

```{r}
data <- data %>% 
  mutate(partyid = fct_lump(partyid, n = 4))

data %>% 
  count(partyid) %>% 
  mutate(share = round(n/sum(n)*100, 2)) %>% 
  select(!n) %>% 
  kable(col.names = c("Party", "% Share")) %>% 
  kable_paper() %>% 
  kable_styling(full_width = F, position = "center")

```

## C. Renaming (lumping)
```{r}
party_share <- data %>% 
               count(partyid) %>% 
               mutate(share = round(n/sum(n)*100, 2))

ggplot(party_share, aes(x = partyid, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = partyid,
                   yend = share,
                   xend = partyid), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Political Parties",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("party affilation") +
  ylab("") +
  scale_y_continuous(limits = c(0,50),
    labels = function(x) paste0(x, "%")) +
  coord_flip()
```

## C. Renaming (lumping) {.build .smaller}
As a review, we reorder the factor based on the share:

`data %>%` 
`mutate(partyid = fct_reorder(partyid, share))`

## C. Renaming (lumping)

```{r}
party_share <- party_share %>% 
               mutate(partyid = fct_reorder(partyid, share))

ggplot(party_share, aes(x = partyid, y = share)) +
  geom_point(color = "#007db7", stat = "identity", size = 6) +
  geom_segment(aes(y = 0,
                   x = partyid,
                   yend = share,
                   xend = partyid), 
               color = "#007db7",
               size = 1.5) +
  charts_theme + 
  labs(title = "Share of Political Parties",
       subtitle = "General Social Survey, 2000-2014") +
  xlab("party affilation") +
  ylab("") +
  scale_y_continuous(limits = c(0,50),
    labels = function(x) paste0(x, "%")) +
  coord_flip()
```

# Summary {data-background=#68c5ea}

## forcats functions {.build .smaller}

For **reordering factors**: \

1. `fct_relevel()` - for reordering factors manually or by hand \
2. `fct_reorder()` - for reordering factors based on another variable \
3. `fct_infreq()` - for reordering factors based on frequency \
4. `fct_rev()` - for reversing the order of factors \

<br>

For **renaming factors**: \

1. `fct_recode()` - for renaming factor levels manually or by hand \
2. `fct_collapse()` - for combining multiple factor levels \
3. `fct_lump()` - for combining most infrequent factor levels

# References {data-background=#68c5ea}

## References

1. Tidyverse forcats tutorial: https://forcats.tidyverse.org \
2. R for Data Science, Chapter 15 Factors: https://r4ds.had.co.nz/factors.html

# Exercises {data-background=#007db7} 
